Global code motion of congruent instructions

### 2017-11-08 cgo/paper.tex
chi args

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules congruent
instructions across the program.  Not only \GCM{} saves code size, it exposes
redundancies in some cases, it exposes more instruction level parallelism in the
basic-block to which instructions are moved, and it enables other passes like
loop invariant motion to remove more redundancies. The cost model to drive the
code motion is based on liveness analysis on \SSA{} representation such that the
(virtual) register pressure does not increase resulting in 2\% fewer spills on
the \SPEC{} benchmark suite.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in the
majority of benchmarks. We have also seen execution time improvements in a few
of \SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by
2\% on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We also formalize why register pressure
reduces as a result of \gcm{}, and how \gcm{} increases instruction level
parallelism thereby enabling more \ooo{} execution on modern architectures.
\end{abstract}

@@

This paper describes \gcm{} (\GCM{}) of congruent instructions
\cite{briggs1997}, a technique that uses the information computed for \PRE{} to
detect identical computations but has a transformation phase whose goal differs
from \PRE{}: it removes identical computations from different branches of
execution.  These identical computations in different branches of execution are
not redundant computations at run-time and the number of run-time computations
is not reduced. It is not a redundancy elimination pass, and thus it has
different cost function and heuristics than \PRE{} or CSE. It is similar to
global code scheduling \cite{dragonbook,click1995global} in the sense that it
will only move computations. Code hoisting, for example, can reduce the critical
path length of execution in \ooo{} machines. As more instructions are available
at the hoisting point, the hardware has more independent instructions to
reorder. The following reduced example illustrates how hoisting can improve
performance by exposing more ILP.

@@

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
seen the performance of the original benchmark (on an \ooo{} processor) improve
by $15\%$.  In fact, there are several advantages of \GCM{}:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism (ILP) to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles. Although it decreases the ILP of basic blocks from where the
  instructions were removed, those basic blocks execute for lesser number of
  cycles than the destination basic block.
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single basic-block, should all the instructions in a conditional get
  hoisted or sunk;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-motion is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP) which results in prohibitively
high compile time which is not suitable for industrial compilers. To the best of
our knowledge we have not found any reference which explored \gcm{} of congruent
instructions. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk. The main contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item an optimistic fast algorithm to schedule congruent instructions;
\item introducing an analogue of PHI nodes on the inverse CFG which allows
  faster tracking of anticipable instructions;
\item a cost model to reduce interference and hence, reduce spills;
\item experimental evaluation of our implementation in \LLVM{}.
\end{itemize}

@@

A CHI keeps information about values flowing out of a basic block.  It is
similar to PHI but in the inverse graph, and used for outgoing values on each
edge. Although it may be suitable for other analyses/optimizations, for \GCM{},
it is computed only for instructions with multiple occurrences in the CFG. An
illustration of CHI is shown in Figure~\ref{fig:chi-intro}. The \GVN{} for both
I1 and I2 is V, the CHI node will save the instruction as well as the edge where
the value is flowing to.  Where CHIArg1, CHIArg2 are of type CHIArg. The CHI for
both I1 and I2 is inserted in their PDF. For sinking we use the similar approach
as \cite{ssapre} to track values flowing into a basic block.

@@

When SSAPRE inserts PHI (expression PHIs), it does so at places where
(potentially) multiple expressions may merge. If a PHI has a missing entry, that
means the expression is partially available.  The concept may work for hoisting
expression to help factor out anticipability by working on inverted graph. If we
insert a CHI, to a basic block with multiple successors, as instructions are
hoisted upwards to the post-dominance frontier.  And we start walking the CFG to
figure out if any CHI has a missing entry, that means the expression is not
anticipable in the basic block having that outgoing PHI.

@@

For sinking we reuse the PHI nodes instead of inserting another expression nodes
like in \cite{ssapre} because PHI nodes ensure availability of definitions, and
hence downward safety, as long as all instructions corresponding to a PHI are
congruent. Also, the PHI nodes represent a complete set of instructions which
could be potentially sunk for code-size optimizations because PHI nodes
represent all values flowing out from a basic block into its dominance-frontier.

\begin{verbatim}
int doCodeSinking()
  // Find sinkable instructions
  For each value number VN > 1 instructions:
    if there are two instructions with same immediate
        successor(S) as post-dominator:
      if both are only used in the same PHI-node of S
          && dependencies of both are available at S:
         mark instructions as sinkable

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules congruent
instructions across the program.  Not only \GCM{} saves code size, it exposes
redundancies as well, it exposes more instruction level parallelism in the
basic-block to which instructions are moved, and it enables other passes like
loop invariant motion to remove redundancies. The cost model to drive the code
motion is based on liveness analysis on \SSA{} representation such that the
(virtual) register pressure does not increase resulting in 2\% fewer spills on
the \SPEC{} benchmark suite.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in the
majority of benchmarks. We have also seen execution time improvements in a few
of \SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by
2\% on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We also formalize why register pressure
reduces, and how \GCM{} increases instruction level parallelism thereby enabling
more \ooo{} execution on modern architectures.
\end{abstract}

@@

This paper describes a fast algorithm for \gcm{} (\GCM{}) of congruent
instructions \cite{briggs1997}, a technique that uses the information computed
for \PRE{} to detect identical computations but has a transformation phase whose
goal differs from \PRE{}: it removes identical computations from different
branches of execution.  These identical computations in different branches of
execution are not redundant computations at run-time and the number of run-time
computations is not reduced. It is not a redundancy elimination pass, and thus
it has different cost function and heuristics than \PRE{} or CSE. It is similar
to global code scheduling \cite{dragonbook,click1995global} in the sense that it
will only move computations. Code hoisting, for example, can reduce the critical
path length of execution in \ooo{} machines. As more instructions are available
at the hoisting point, the hardware has more independent instructions to
reorder. The following reduced example illustrates how hoisting can improve
performance by exposing more ILP.

@@

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}.  Although
it decreases the instruction level parallelism (ILP) of basic blocks from where
the instructions were removed, those basic blocks cannot execute for more cycles
than the destination basic block. We have seen the performance of a proprietary
benchmark (on an \ooo{} processor) improve by $15\%$.  In fact, there are
several advantages of \GCM{}:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item it helps reduce the code size of the program by replacing multiple
  instructions with one thereby making the function cheaper to inline because
  inliner heuristics depend on instruction count;
\item it exposes more ILP to the later compiler passes. By hoisting identical
  computations to be executed earlier, instruction schedulers can move heavy
  computations earlier in order to avoid pipeline bubbles.
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication \cite{simd};
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single basic-block, should all the instructions in a conditional get
  hoisted or sunk;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-motion is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There has been a lot of work both in industry and academia related to code
hoisting/sinking, and in general global scheduling \cite{click1995global}. Some
relate code-hoisting to code-size optimization \cite{rosen1988global} and many
\cite{barany2013, shobaki2013} use global scheduling to improve
performance. Most of the recent work on global scheduling are done using integer
linear programming (ILP) which results in prohibitively high compile time which
is not suitable for industrial compilers. To the best of our knowledge we have
not found any reference which explored \gcm{} with a cost-model to reduce
register-pressure. A part of our implementation (aggressive code hoisting) is
already merged in \LLVM{} trunk. The main contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item an optimistic fast algorithm to schedule congruent instructions;
\item introducing an analogue of $\phi$ nodes on the inverse CFG which allows
  faster tracking of anticipable instructions;
\item a cost model to reduce interference and hence, reduce spills;
\item experimental evaluation of our implementation in \LLVM{}.
\end{itemize}

@@

A CHI keeps information about values flowing out of a basic block.  It is
similar to $\phi$ \cite{ssa} but in the inverse graph, and used for out-flowing
values on each edge. Although it may be suitable for other
analyses/optimizations, for \GCM{}, it is computed only for instructions with
multiple occurrences in the CFG. An illustration of CHI is shown in
Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the CHI node
will save the instruction as well as the edge where the value is flowing to.
Where CHIArg1, CHIArg2 are of type CHIArg. The CHI for both I1 and I2 is
inserted in their PDF (A). For sinking we use the similar approach as
\cite{ssapre} to track values flowing into a basic block.

@@

When SSAPRE inserts $\phi$ (expression PHIs), it does so at places where
(potentially) multiple expressions may merge. If a $\phi$ has a missing entry, that
means the expression is partially available.  The concept may work for hoisting
expression to help factor out anticipability by working on inverted graph. If we
insert a CHI, to a basic block with multiple successors, as instructions are
hoisted upwards to the post-dominance frontier.  And we start walking the CFG to
figure out if any CHI has a missing entry, that means the expression is not
anticipable in the basic block having CHI.

@@

For sinking we reuse the $\phi$ nodes instead of inserting another expression nodes
like in \cite{ssapre} because $\phi$ nodes ensure availability of definitions, and
hence downward safety, as long as all instructions corresponding to a $\phi$ are
congruent. Also, the $\phi$ nodes represent a complete set of instructions which
could be potentially sunk for code-size optimizations because $\phi$ nodes
represent all values flowing out from a basic block into its dominance-frontier.

\begin{verbatim}
int doCodeSinking()
  // Find sinkable instructions
  For each value number VN > 1 instructions:
    if there are two instructions with same immediate
        successor(S) as post-dominator:
      if both are only used in the same $\phi$-node of S
          && dependencies of both are available at S:
         mark instructions as sinkable
